---
title: "Multidimensional Statistical Analysis"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
---

Lamyaa BOUZBIBA
Alexandre POUPEAU
Eloise JULIEN

```{r}
NAm2 = read.table("NAm2.txt", header = TRUE)
#NAm2

cont <- function(x)
{ if (x %in% c("Canada"))
  cont <-"NorthAmerica"
  else if (x %in% c("Guatemala","Mexico","Panama","CostaRica"))
  cont <-"CentralAmerica"
  else
  cont<-"SouthAmerica"
  return (factor(cont))
}
contID <- sapply(as.character(NAm2[,4]),FUN=cont)

library("class")
library("MASS")

```


#### a) 

```{r}

labels <- rep (1:2, each = 494/2)
set = sample(labels, 494)

# continent + genetic markers
NAcont <- cbind(contID=contID,NAm2[,-(1:8)])

NAcont[,1] <- factor(NAcont[,1])

# we make our training phase using the elements which have its set value equal 1.
lda(contID ~ . , data = NAcont[, 1:1001], subset = (set == 1))

```

Here, lda does not work because there are genetic markers with a variance of 0 so the intra_group covariance matrix is non-reversing and R cannot solve it. Actually, collinear variables should be removed.


#### b)

```{r}

withinvar <- apply(NAcont[(set==1),-1],FUN=function(x){tapply(x,NAcont[(set==1),1],FUN=var)},MARGIN=2)

bool <- as.logical(apply(withinvar,FUN=function(x){prod(x!=0)},MARGIN=2))

NAcont2 <- cbind(contID=contID,(NAm2[,-(1:8)])[,bool])

NAcont2[,1]<-factor(NAcont[,1])

classModel <- lda(contID ~ . , data = subset(NAcont2, set == 1))

#classModel
```

Instead of $5717$, there are $3136$ genetic markers remaining after removing those which variance is equal to zero. 

#### c)

Now we predict the original populations of individuals who have not been trained.

```{r}

# Here we make our test phase using the elements which have its set value equal 2.
prediction <- predict(classModel, newdata = subset(NAcont2, set == 2))
#prediction$class

```

#### d)

```{r}

confusionMatrix <- table(subset(contID, set == 2), prediction$class)
confusionMatrix
barplot(confusionMatrix, col=c("bisque", "bisque3", "bisque4"))
legend("topleft", legend=c("NorthAmerica", "CentralAmerica", "SouthAmerica"), fill=c("bisque", "bisque3", "bisque4"))

```

With this barplot, we can see for instance that we predicted that 1 person is North American instead of South American. Moreover, we predicted that 55 people are South American instead of Central American, this the highest error.


#### e)

```{r}

rate <- 1-((sum(diag(confusionMatrix)))/sum(confusionMatrix))

sprintf("The rate of misclassified individuals is equal to %s .", rate)


```

The classification error is equal to about 30% .


#### f)

A possible improvement to the method implemented in the previous question could be the 13-fold cross-validation. So, we will divide randomly the data into 13 subsets where 12 subsets are used as the training sets and the remaining subset is used as a validation subset to compute a prediction error. We repeat this procedure ten times by considering each subset as the validation subset and the other subsets as the training sets.


```{r}

labels <- rep(1:13, each=13)
set = sample(labels, 494, replace=TRUE)

NAcont <- cbind(contID=contID,NAm2[,-(1:8)])

NAcont[,1] <- factor(NAcont[,1])

rate <- c(1:13)

for (i in 1:13){
  
  withinvar <- apply(NAcont[(set==i),-1],FUN=function(x){tapply(x,NAcont[(set==i),1],FUN=var)},MARGIN=2)

  bool <- as.logical(apply(withinvar,FUN=function(x){prod(x!=0)},MARGIN=2))

  NAcont2 <- cbind(contID=contID,(NAm2[,-(1:8)])[,bool])

  NAcont2[,1]<-factor(NAcont[,1])

  classModel <- lda(contID ~ . , data = subset(NAcont2, set != i))
  
  prediction <- predict(classModel, newdata = subset(NAcont2, set == i))
  
  confusionMatrix <- table(subset(contID, set == i), prediction$class)

  rate[i] <- 1-((sum(diag(confusionMatrix)))/sum(confusionMatrix))
  
}

sprintf("The rate of misclassified individuals is equal to %s .", mean(rate))

```

With the 13-fold cross validation, the classification error is equal to about 18% .

To conclude, this cross validation is a good improvement to the method implemented in the previous question. Indeed, the classification error is equal to about 18% against 30% previously. 







