---
title: "Multidimensional Statistical Analysis"
output: html_notebook
---

```{r}
NAm2 = read.table("NAm2.txt", header = TRUE)
#NAm2

cont <- function(x)
{ if (x %in% c("Canada"))
  cont <-"NorthAmerica"
  else if (x %in% c("Guatemala","Mexico","Panama","CostaRica"))
  cont <-"CentralAmerica"
  else
  cont<-"SouthAmerica"
  return (factor(cont))
}
contID <- sapply(as.character(NAm2[,4]),FUN=cont)

library("class")
library("MASS")

```


#### a) 

```{r}

labels <- rep (1:2, each = 494/2)
set = sample(labels, 494)

# continent + genetic markers
NAcont <- cbind(contID=contID,NAm2[,-(1:8)])

NAcont[,1] <- factor(NAcont[,1])

# we make our training phase using the elements which have its set value equal 1.
lda(contID ~ . , data = NAcont[, 1:1001], subset = (set == 1))


```

Here, lda does not work because there are genetic markers with a variance of 0 so the intra_group covariance matrix is non-reversing (determinant=0) and R cannot solve it. Actually, collinear variables should be removed.


#### b)

```{r}

withinvar <- apply(NAcont[(set==1),-1],FUN=function(x){tapply(x,NAcont[(set==1),1],FUN=var)},MARGIN=2)

bool <- as.logical(apply(withinvar,FUN=function(x){prod(x!=0)},MARGIN=2))

NAcont2 <- cbind(contID=contID,(NAm2[,-(1:8)])[,bool])

NAcont2[,1]<-factor(NAcont[,1])

classModel <- lda(contID ~ . , data = subset(NAcont2, set == 1))

#classModel
```

There are $3136$ genetic markers instead of $5717$ remaining after removing those which variance is equal to zero. 

#### c)

```{r}

# Here we make our test phase using the elements which have its set value equal 2.
prediction <- predict(classModel, newdata = subset(NAcont2, set == 2))
#prediction$class

```

#### d)

```{r}

confusionMatrix <- table(subset(contID, set == 2), prediction$class)
confusionMatrix
barplot(confusionMatrix, col=c("bisque", "bisque3", "bisque4"))
legend("topleft", legend=c("NorthAmerica", "CentralAmerica", "SouthAmerica"), fill=c("bisque", "bisque3", "bisque4"))

```

With this barplot, we can see for instance that we predicted that 1 person is North American instead of South American. Moreover, we predicted that 38 people are South American instead of Central American, this the highest error. In the other cases, the rate of error is weak in comparison to what it has been well predicted.


#### e)

```{r}

sum <- 0
for (i in 1:nrow(confusionMatrix)) {
  for (j in 1:ncol(confusionMatrix)) {
    if (i != j) {
      sum <- sum + confusionMatrix[i, j]
    }
  }
}

sprintf("The rate of misclassified individuals is equal to %s .", sum/(length(NAcont)/2))

```

The classification error is equal to about 2.1% .


#### f)

```{r}

```










